---
pagetitle: 'Population Dynamics - Solutions'
output: 
  html_document:
    toc: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy(position = c('top', 'right'))
```

# Exercise: Population Dynamics {.tabset}

## Question 1
You are studying a closed population of beetles consisting of 3000 individuals. You register 400 births and 150 deaths during one month. Calculate *r* and *lambda* and predict the population size after 6 months.

**Answer:** 
We assume 400 births and 150 deaths within a month. Build a model based on a projection interval of one month. Then
$$
\begin{align}
N_{t+1} &= N_t + B - D = 3000 + 400 - 150 = 3250\\
\lambda &= \frac{N_{t+1}}{N_{t}} = \frac{3250}{3000} = 1.0833\\
r &= \ln \lambda = \ln(1.0833) = 0.08\\
N_6 &= 3000* \exp(0.08*6) = 3000*1.0833^6= 4848
\end{align}
$$
Thus, after six month we expect 4848 beetles.

## Question 2
You estimate the size of a flatworm population to be the following at five consecutive days: 100, 158, 315, 398 and 794 individuals. Plot the natural logarithm (i.e., to the base e) of the population size and estimate *r*. You can plot the data by hand on a piece of paper, or by using EXCEL or R.

**Answer:**
Start by taking the natural logarithm of the five population estimates, resulting in 4.605, 5.063, 5.753, 5.986, and 6.677.
Plot these as a function of time (days).

Draw a straight line through the data points. There are exact mathematical methods, for example, the method of least squares (linear regression) that can be executed in R or Excel. Here you can see the R-code for doing this.

```{r}
x <- 1:5
y <- c(4.605, 5.063, 5.753, 5.986, 6.677)

plot(x, y, xlab = "day", ylab = "population size")
linear.regression <- lm(y ~ x)
linear.regression
abline(linear.regression)
```

The regression analysis also shows the value for the slope, namely $0.5067$. This slope is equal to the population growth rate $r$.

For the purpose of this exercise it is sufficient to draw the line with the "naked eye" and estimate its slope by dividing the change in $y$ by the change in $x$, $\Delta y/\Delta x$.

## Question 3
An isolated population of elk (a species of deer) in Washington state increased from 1975 to 1993 as follows:

$$
\begin{align*}
    1975 &\,\,\,\,\, 8  & 1986 &\,\,\,\,\, 68 \\
    1976 &\,\,\,\,\, 13 & 1987 &\,\,\,\,\, 67 \\
    1977 &\,\,\,\,\, 14 & 1988 &\,\,\,\,\, 72 \\
    1979 &\,\,\,\,\, 15 & 1989 &\,\,\,\,\, 79 \\
    1981 &\,\,\,\,\, 20 & 1990 &\,\,\,\,\, 94 \\
    1982 &\,\,\,\,\, 21 & 1991 &\,\,\,\,\, 110 \\
    1983 &\,\,\,\,\, 27 & 1992 &\,\,\,\,\, 146 \\
    1984 &\,\,\,\,\, 40 & 1993 &\,\,\,\,\, 179 \\
    1985 &\,\,\,\,\, 54 &&
\end{align*}
$$
Plot these data.
What shape of population growth is shown by these data? Could one fit the logistic equation to these data? Use a discrete-time model to calculate the net reproductive ratio $\lambda$ for this population.

**Answer:**
```{r}
x <- c(1975, 1976, 1977, 1978, 1979, 1980, 1981, 1982, 1983, 1984, 1985, 1986, 1987, 1988, 1989, 1990, 1991, 1992, 1993)
y <- c(8, 13, 14, NA, 15, NA, 20, 21, 27, 40, 54, 68, 67, 72, 79, 94, 110, 146, 179)

plot(x, y, xlab = "year", ylab = "population size")
```
There is no evidence for the curve to level off around an equilibrium *K*. Rather, the elk population seems to grow exponentially.

We can use the same method as under Q2 to calculate $\lambda$:
```{r}
lny <- log(y)
plot(x, lny, xlab = "day", ylab = "population size")
linear.regression <- lm(lny ~ x)
linear.regression
abline(linear.regression)
```

Thus $r=0.1635$, and
$$
\lambda=\exp(r)=\exp(0.1635)=1.177625.
$$

Alternatively, solving for *r* can done as follows:
$$
\begin{align}
N_t &= N_0 \lambda^{t}\\
\lambda^{t} &= \frac{N_t}{N_0}\\
\lambda &=\left(\frac{N_t}{N_0}\right)^{\frac{1}{t}}=\sqrt[^{t}]{\frac{N_t}{N_0}}\\
\lambda &= \sqrt[^{18}]{\frac{179}{8}} = \sqrt[^{18}]{22.375}\approx1.19
\end{align}
$$

## Question 4
Suppose that a butterfly population shows logistic growth with *K = 500* and *r = 0.1* individuals/(individuals*month). What is the maximal growth rate for this population?

**Answer:**
Recall that the growth rate of a logistically growing population is highest when population size equals *N = K/2*. Since *K = 500* this population will grow fastest at *N = 500/2 = 250*. Then
$$
\frac{\mathrm{d}N}{\mathrm{d}t} = rN\left(1-\frac{N}{K}\right) \Longrightarrow 0.1*250\left(1-\frac{250}{500}\right)= 12.5.
$$
Thus, the maximal population growth rate equals 12.5 individuals per month.

## Question 5
A fisher is maximizing their fishing yield by maintaining a population of logistically growing lake trout at the density where the population growth rate is fastest. For the population in question, this occurs at a density of 500 individuals. Predict the realized population growth rate after the population has been stocked with an additional 600 fish. Assume that for the trout *r=0.005* individuals/(individual*day).

**Answer:**
The fact that this logistically growing population is growing fastest at *K/2=500* individuals implies that *K = 1000*. With 600 fish added the new population size equals 1100. Then
$$
\frac{\mathrm{d}N}{\mathrm{d}t} = rN\left(1-\frac{N}{K}\right) \Longrightarrow 0.005*1100\left(1-\frac{1100}{1000}\right) = -0.55.
$$
Thus, the growth rate equals -0.55 fish per day.
The growth rate is negative because the additional stock pushes the population above its carrying capacity.
