---
pagetitle: 'Exercise: Population Dynamics'
output: 
  html_document:
    toc: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy(position = c('top', 'right'))
```

# Exercise: Population Dynamics {.tabset}

## Question 1
You are studying a closed population of beetles consisting of 3000 individuals. You register 400 births and 150 deaths during one month. Calculate *r* and *lambda* and predict the population size after 6 months.

## Question 2
You estimate the size of a flatworm population to be the following at five consecutive days: 100, 158, 315, 398 and 794 individuals. Plot the natural logarithm (i.e., to the base e, where e is Eulers number) of the population size and estimate *r*. You can plot the data by hand on a piece of paper, or by using EXCEL or R.

## Question 3
An isolated population of elk (a species of deer) in Washington state increased from 1975 to 1993 as follows:

$$
\begin{align*}
    1975 &\,\,\,\,\, 8  & 1986 &\,\,\,\,\, 68 \\
    1976 &\,\,\,\,\, 13 & 1987 &\,\,\,\,\, 67 \\
    1977 &\,\,\,\,\, 14 & 1988 &\,\,\,\,\, 72 \\
    1979 &\,\,\,\,\, 15 & 1989 &\,\,\,\,\, 79 \\
    1981 &\,\,\,\,\, 20 & 1990 &\,\,\,\,\, 94 \\
    1982 &\,\,\,\,\, 21 & 1991 &\,\,\,\,\, 110 \\
    1983 &\,\,\,\,\, 27 & 1992 &\,\,\,\,\, 146 \\
    1984 &\,\,\,\,\, 40 & 1993 &\,\,\,\,\, 179 \\
    1985 &\,\,\,\,\, 54 &&
\end{align*}
$$

Plot these data as function of time. What shape of population growth is shown by these data? Could one fit the logistic equation to these data? Use a discrete-time model to calculate the net reproductive ratio *lambda* for this population.

## Question 4
Suppose that a butterfly population shows logistic growth with *K = 500* and *r = 0.1* individuals/(individuals*month). What is the maximal population growth rate for this population?

## Question 5
A fisher is maximizing their fishing yield by maintaining a population of logistically growing lake trout at the density where the population growth rate is fastest. For the population in question this occurs at a density of 500 individuals. Predict the realized population growth rate after the population has been stocked with an additional 600 fish. Assume that for the trout $r=0.005$ individuals/(individual*day).
